---
title: "data_analysis_group4"
authors: "Razak Diallo, Lea Glaser, Deja Jones, Alexander Taylor"
date: "2025-12-06"
output: html_document
---

# Load Libraries
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(tidycensus)
```


# Load and clean data
```{r}


state_payments <- read_csv("data/state_payments.csv") |>
  clean_names() |>
  mutate(
    date = ymd_hms(date),
    amount = str_replace_all(amount, "\\$", ""),
    amount = str_replace_all(amount, "\\,", ""),
    amount = as.numeric(amount)
         )

glimpse(state_payments)


```
# Question 1: What is the median income of the zip codes where the top five vendors are located?
# Findings: 
```{r}

```

# Question 2: What year did the State of Maryland spend the most? How about the least?
# Findings: 
```{r}

```

# Question 3: What is the total amount of payments in each zip code?
# Findings: 
```{r}

```

# Question 4: How many payments were made each month?
# Findings: 
```{r}

```

# Question 5: What agency spent the most money in each county?
# Findings: 
```{r}

```
